---
title: "Untitled"
author: "Me"
date: "September 30, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a quick post intended for animating how the transition matrix of a Markov chain changes between larger time steps, as well as showing the probability of the chain being in any specified state over time.  This post uses the tidyverse, along with `gganimate`.

```{r, message=FALSE}
library(tidyverse)
library(magrittr)  ## using some aliases not loaded by default
library(gganimate)
```


## The Transition Matrix

The transition matrix for a Markov chain describes the probabilities of the state moving between any two values; since Markov chains are memoryless, these probabilities hold for all time steps.  It is a square matrix like this:

\[ M =\begin{bmatrix} 0.7 & 0.2 & 0.1 \\ 0.2 & 0.5 & 0.3 \\ 0 & 0 & 1 \end{bmatrix}\]

where the row indicates the state you're starting in and the column indicates where you would end up.  Since these are probabilities, each row adds up to 1.  Values range from 0 to 1 inclusive; a row with a 1 on the matrix's diagonal is an absorbing state, meaning that once the Markov chain is in that state, it will never leave it.

A useful property of the transition matrix is that it can represent the long-term probability of ending up in one state given a starting state by just taking a large exponent of the matrix itself.  And since it is just a matrix, it's easy to visualize it with a heatmap, so viewing how those probabilities shift over time is fairly simple.

Similarly, the overall probability of the system being in a state at any given time can be seen over time.  If the transition matrix $M$ converges to a constant value, then if we have a starting probability vector $s_0$, there should some future point $t$ where $s_{t+1} \approx s_t M$ for all steps starting at $t$.

## The Functions

The intent is to animate the transition matrix with R's `gganimate` library.  However, `gganimate` requires `ggplot`, which expects the matrices to be in a long format.  So a function to transform it appropriately is needed.

```{r}
## Turns a matrix into a three column data frame: row of the cell, column of the
## cell, and the value.
matrix_to_long <- function(m){
    m_long <- m %>%
        set_rownames(1:dim(m)[1]) %>% set_colnames(1:dim(m)[2]) %>%
        as.data.frame() %>%
        rownames_to_column("Row") %>%
        pivot_longer(-c(Row), names_to="Column", values_to="P") %>%
        mutate(Row = as.numeric(Row), Column = as.numeric(Column))
    return(m_long)
}
```

Next, we need functions for calculating the values of the transition matrix and the probability vector at each time step.

```{r}
## s, m, and t are the starting probability vector, transition matrix, and 
## number of time steps respectively

transition_matrix_over_time <- function(m, t){
    m_list <- lapply(1:t, function(i){m})
    m_at_steps <- Reduce(`%*%`, m_list, accumulate = TRUE)
    m_at_steps <- append(m_at_steps, list(m), 0)
    return(m_at_steps)
}

probability_vector_over_time <- function(s, m, t){
    transition_matrices <- transition_matrix_over_time(m, t)
    probability_vectors <- lapply(transition_matrices, function(mtrx){s %*% mtrx})
    probability_vectors <- append(probability_vectors, list(c(s)), 0)
    return(probability_vectors)
}
```

The `Reduce` function is the key to keeping this relatively compact.  It iteratively computes the result of the binary function on the first two arguments, then on that first result and the third element, and so on.  Using `accumulate=TRUE` ensures that every intermediate result is returned from the function, and not just the final result.

Finally, we need to return the plots.  First, the animated transition matrix plot:

```{r}
transition_matrix_animation <- function(m, t){
    m_over_time <- transition_matrix_over_time(m, t)
    long_data <- lapply(1:length(m_over_time),
                        function(x){data.frame(N=x-1, matrix_to_long(m_over_time[[x]]))})
    long_data <- do.call("rbind", long_data)
    g <- ggplot(long_data, aes(x=Column, y=Row, fill=P)) + geom_raster() + 
        geom_text(aes(label=round(P, 2))) + 
        scale_y_reverse() + transition_states(N) +
        ggtitle("Transition matrix at time t={closest_state}")
    return(g)
}
```

Note that the vertical axis is being reversed.  This is a convenience, so that the first row of the transition matrix is the top row in the plot -- since the `Column` and `Row` values are both numeric, R would have the matrix upside-down otherwise.

Then the plot for the states' probabilities.  Though doing this as a one-dimensional heatmap was certainly possible, a line chart is probably still more effective for showing change over time.

```{r}
state_probability_plot <- function(s, m, t){
    state_probabilities <- probability_vector_over_time(s, m, t)
    state_prob_matrix <- do.call("rbind", state_probabilities)
    state_prob_df <- state_prob_matrix %>% as.data.frame() %>%
        set_colnames(c("State1", "State2", "State3", "State4")) %>%
        rownames_to_column("t") %>%
        mutate(t = as.integer(t)-1) %>%
        pivot_longer(cols=-c(t), names_to="State")
    g <- ggplot(state_prob_df) + geom_line(aes(x=t, y=value, color=State)) +
        ylim(c(0, max(state_prob_df$value)+0.05))
    return(g)
}
```


## Examples

I'll starting with the following transition matrix:

```{r}
m1 <- matrix(c(0.8,  0.1,  0.1,  0,
               0.05, 0.85, 0,    0.1,
               0.05, 0,    0.9,  0.05,
               0.3,  0.2,  0.1,  0.4),
             byrow=TRUE, nrow=4)
s1 <- c(0.25, 0.25, 0.25, 0.25)
```
There are no absorbing states in this matrix, though there are a few impossible transitions.  Additionally, the first three states are likely to stay in the same state if it's already there, but the fourth state is more likely than not to transition to something else.


```{r}
g1 <- transition_matrix_animation(m1, 50)
anim_save("m1_anim.gif", g1, nframes=102)

state_probability_plot(s1, m1, 50)
```


Let's move on to considering a chain with one absorbing state.

```{r}
m2 <- matrix(c(0.6,  0.29, 0.09, 0.02,
               0.14, 0.8,  0.05, 0.01,
               0.08, 0.5,  0.4,  0.02,
               0,    0,    0,    1),
             byrow=TRUE, nrow=4)
s2 <- c(0.25, 0.25, 0.25, 0.25)
```

```{r}
g2 <- transition_matrix_animation(m2, 50)
anim_save("m2_anim.gif", g2, nframes=102)

state_probability_plot(s2, m2, 50)
```

Since everything eventually converges to the one absorbing state in this case, we would expect it to eventually end up like this -- the absorbing state becomes increasingly inevitable as time goes on.

How about two absorbing states?

```{r}
m3 <- matrix(c(1,    0,    0,    0,
               0.14, 0.8,  0.05, 0.01,
               0.08, 0.5,  0.4,  0.02,
               0,    0,    0,    1),
             byrow=TRUE, nrow=4)
s3 <- c(0, 0.5, 0.5, 0)
```

```{r}
g3 <- transition_matrix_animation(m3, 50)
anim_save("m3_anim.gif", g3, nframes=102)

state_probability_plot(s3, m3, 50)
```

This is a bit more interesting than the previous one, since things are going to be split between the two absorbing states.

(note: plotly could also be good)